Probabilistic bioregion models suggest seabird assemblages are linked to the major boundary current off eastern Australia

### 2023-07-10 docs/ms-draft.Rmd
review from Will, plus partial plots sentences

---
title: "Probabilistic bioregion models suggest seabird assemblages are linked to the major boundary current off eastern Australia"
author:
  - Nicholas W. Daudt:
      email: nicholaswdaudt@gmail.com
      institute: [MarSciUO]
      correspondence: true
  - Eric J. Woehler:
      institute: [ASG]
      correspondence: false
  - Matthew R. Schofield:
      institute: [MathStatUO]
      correspondence: false
  - Robert O. Smith:
      institute: [MarSciUO]
      correspondence: false
  - Leandro Bugoni:
      institute: [FURG]
      correspondence: false
  - William J. Rayment:
      institute: [MarSciUO]
      correspondence: false
institute:
  - MarSciUO: Department of Marine Science, University of Otago, Aotearoa New Zealand
  - ASG: Australasian Seabird Group, BirdLife Australia, Hobart, Australia
  - MathStatUO: Department of Mathematics and Statistics, University of Otago, Aotearoa New Zealand
  - FURG: Institute of Biological Sciences, Universidade Federal do Rio Grande--FURG, Rio Grande, RS, Brazil
abstract: |
  **Aim**: To identify assemblages of seabird species at large scale, probabilistically, and related them to environmental conditions off eastern Australia, a world seabird hotspot.
  **Methods**: We used seabird at-sea ship-based survey data collected over 37° latitude and fitted Region of Common Profile (RCP) mixture models for each season and type of data (presence-absence and counts). RCP groups are defined as a set of regions where the probability of encountering a specific species profile is constant within it but different from other regions. These groups also vary according to covariates, on which we included oceanographic static and climatic, as well as physiographic layers.
  **Results**: All models suggested two large scale assemblages ('northern' and 'southern' assemblages), except for the abundance-autumn model which indicated three groups. Model results seem to agree in a biogeographic barrier at ~32°S, where the East Australian Current mostly dissipates to the east and loses force. Furthermore, sea surface temperature and, in one case, sea surface salinity, were selected in all final models, further indicating close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and counts resulted in similar spatial and species profile patterns.
  **Conclusion**: Two seabird assemblages were clearly identified from RCP models, suggesting the persistence of these groups at temporal and spatial scales. Given the rapid change that the East Australian Current is experiencing, and its importance on driving seabirds' distributions, the methods applied in our study could be replicated in future analyses to assess possible spatial/temporal changes in seabird assemblages, including on its structure.
output: 
  # rticles::sim_article:
  #   keep_tex: TRUE
  bookdown::word_document2:
    pandoc_args:
      - --lua-filter=../docs/scholarly-metadata.lua
      - --lua-filter=../docs/author-info-blocks.lua
      - --reference-doc=../docs/docx-template.docx 
bibliography: ../references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
---

@@

<!-- 'rticles::sim_article' example? -->

<!-- https://github.com/emitanaka/combine2019/tree/master/output/articles/article_template_sim_article -->

```{r setup, include = FALSE, echo = FALSE}
library(gt)
```

@@

Bioregionalization is the discipline that identifies, delimits, characterises, and names biogeographical regions [@vilhena2015; @morrone2018]. Although there is no agreement on the naming systems between proposed regionalizations [but see @ebach2008], the hierarchical nature of bioregions is well recognized [@vilhena2015; @kreft2010]. Thus, bioregions are usually described within geographically larger classifications, such as 'provinces' or 'realms' [@spalding2007; @costello2017]. The delimitation of bioregions is made through identifying sets of species that are endemic or with high levels of co-occurrence, distinct enough from other regions, which is interpreted as a natural representation of common historical and/or ecological processes shaped by physical and biological forces [@kreft2010]. In particular, the identification of bioregions allows the representation of the distribution of multiple species at a given time and space as well as to relate these regions to environmental characteristics. By doing so, bioregions can be interpreted as species assemblages [@lyons2017; @hill2017]. Given these features, bioregions have a fundamental role in conservation biogeography [@whittaker2005], by being used as management units and even shedding light into conservation gaps [@spalding2007].

Seabird assemblages have been described and related to environmental characteristics worldwide. Considering the hierarchical nature of oceanographic processes and its resulting features [@haury1978patterns], it is not surprising that the distribution of high mobile seabirds are linked differently across scales [@hunt1987]. In the marine environment, large scale oceanographic features (water masses and currents) act as 'biomes' and influence seabird assemblage structure--i.e. the occurrence of species and its relative numerical composition [@ribic1997; @hyrenbach2007; @commins2014]. On the other hand, at smaller scales, local abundance is related to temporary processes that aggregate prey and therefore predators, such as fronts and eddies [@weimerskirch2007; @scales2014]. Such smaller-scale processes, although ephemeral, are primarily driven by physical and atmospheric forces and are thus somewhat predictable in space and time, providing consistent feeding areas for marine predators when they occur [@davoren2013; @hazen2013]. Noticeably, sea surface temperature is a key driver of distribution of many seabird species, at individual- and assemblage-levels, as well as at large (water masses) and smaller (fronts) scales [@dunlop1988; @ribic1997; @quillfeldt2015]. Nonetheless, many regions still lack information regarding multiple seabird species distribution and how they assemble in space and time.

One such area that could benefit from better understanding of seabird distributions is eastern Australia, considered a world seabird hotspot [@karpouzi2007; @ramírez2017]. Surprisingly, however, there are not many studies on seabird distribution relative to the region's relevance for this group [@mott2018; @bernard2021]. Off northwest Australia, at-sea surveys, multivariate analyses [@dunlop1988] and ensemble species distribution models [@lavers2014] of species assemblages have shown that preferential habitat use of tropical species are related to sea surface temperature and salinity, as well as indicated the importance of the continental shelf for some species. In particular, @dunlop1988 suggests that salinity is the main variable influencing species distribution in oceanic waters whereas temperature would be the key factor on the continental shelf. On the other hand, off eastern Australia, there is a scarcity of studies at larger scales. @reid2000 and @evans2021a analysed seabird assemblages and their relationship with oceanography and prey data, however at coarse scale (tens of km), off southeast Tasmania. At a multi-species, assemblage-level, the latter studies seem to be the only ones on eastern Australia; other studies were mostly descriptive species accounts at local scale [@norman1992distribution; @blaber1994distribution], or focused on a single species [@priddel2014].

The East Australian Current (EAC) is the dominant current in eastern Australia, the major western boundary current in the South Pacific [@ganachaud2014]. The EAC has been extensively studied by physical oceanographers, although its link between oceanography and biological processes is still in its infancy. The dynamic features provided by EAC, such as eddies and upwellings, can likely influence marine biota [@suthers2011; @scales2014]. While most of the biological studies done within this system were focused on plankton entrapment within the cores of eddies [@garcia2022; @firme2023; @hassler2011], there is a growing body of work focusing on fisheries [@hobday2010; @young2011; @brieva2015]. These studies emphasize how the assemblage structure and probability of occurrence of a given taxa is driven by EAC. In particular, the potential biological consequence triggered by climate changes is the tropicalization of the biota, given the EAC is strengthening and getting warmer [@cai2005; @poloczanska2007; @hobday2010; @phillips2022], as shown elsewhere [@hastings2020]. Notwithstanding, the relationship between oceanography and marine megafauna along the EAC is poorly known [@suthers2011], albeit they are considered environmental sentinels [@hazen2019]. Some progress has been made using shark and ray movement data [@jaine2014; @niella2020; @niella2022], and satellite tracking have also demonstrated that increased sea surface temperature in EAC decreases penguin foraging success [@carroll2016]. However, information at the assemblage-level is still scarce [but see @evans2021a].

Given east Australia's importance for seabirds and the recent attention on how climate changes in the EAC can affect marine biota [@poloczanska2007], an assessment of seabird spatial distribution at large scale is timely. Seabirds can serve as sentinels for tracking EAC ecological changes in many facets [e.g. @sydeman2012], including by their distribution. Surprisingly, though, seabird distribution is under-represented in climate change studies, despite being pointed as a priority topic [@grémillet2009; @sydeman2012]. In this study, we aim to identify assemblages of seabird species probabilistically and related them to environmental conditions, by using at-sea surveys data and a state-of-the-art modelling technique. We used the mixture-model Region of Common Profiles [RCP; @foster2013], which is under the best current available tools for describing bioregions [@hill2020; @woolley2020]. RCP is model-based approach, which include a series of advantages over traditional algorithm-based statistics [@warton2015; @hill2020]. East Australia supports a high diversity of seabirds and offers wide environmental gradients, from the tropics to the subantarctics. Therefore, we expect to find different assemblages relating to different environmental conditions. In addition, we provide a repeatable framework on which future studies can build up to assess how EAC changes will influence seabird assemblage distributions.

@@

The EAC originates from the South Equatorial Current (SEC), a tropical, nutrient-poor water mass. When the SEC's North Caledonian Jet hits the Queensland Plateau (\~18°S), it splits into a north branch running towards the Gulf of Papua and a south branch, which marks the start of the EAC [@kessler2013; @ganachaud2014]. The South Caledonian Jet completes the EAC system at \~22°S after encountering the shelf break in the southern portion of the Great Barrier Reef (**Fig. 1a**; @kessler2013; @ganachaud2014). The EAC interacts with topographical features on the seafloor, from which oceanographic features and processes arise, such as smaller currents, upwellings, and persistent dynamic eddies [@oke2000; @ridgway2003a; @oke2019]. The EAC is mainly attached to the shelf break, but at \~32°S, it spawns an area of eddy generation that dissipates throughout the western Tasman Sea [@mata2006; @oke2019], the 'EAC eastern extension' (@oke2019a; typically mentioned as the 'Tasman Front'). Although most EAC energy is converted into eddies, the EAC still runs southwards following the shelf break towards Tasmania (the 'EAC southern extension'; **Fig. 1a**), possibly carrying eddy-like features as a transient current [@oke2019].

@@

Seabird data were gathered aboard RV Investigator and RV Falkor during 15 voyages, between 2016--2021, over \~37° of latitude (**Fig. 1b**; **Table S1**). One observer continuously recorded seabirds within a 300-m forward quadrant (eye height **18-m, Eric?** a.s.l) throughout daylight hours while the ship was underway at a constant speed. Species, the number of birds and behaviour were logged directly onto a data entry system through a ship-based web server. Individual records received a date/time/geographic coordinate label (multi-species records have the same label but are logged as separate entries for each species) **(Eric: is there any ref for the protocol?)**.

```{r fig1-study-area, fig.cap = "**FIGURE 1.** Study area showing the main ocean currents (a), and seabird records made by season on top of the 1° latitude × 1° longitude grid cells (b). In (a), the East Australian Current (EAC) system is highlighted with its acronyms in bold and drawn in blue color. The EAC system originates from the South Equatorial Current (SEC) and its branches, North Caledonian Jet (NCJ) and South Caledonian Jet (SCJ), drawn in red. The EAC 'eastern extension' is an eddy field (blue shaded area) often mentioned as the 'Tasman Front', from which the East Auckland Current originates (EAUC, in green). Figure (a) is adapted after Oke, Roughan, et al. (2019); for colour references see the online version.", fig.align = "center", echo = FALSE}

@@

To explain and predict species distribution and biogeography, a thoughtful choice of covariates is important [@barbet-massin2014]. We used seven oceanographic variables and calculated three climatic layers from them, and three physiographical variables (**Table 1**). These variables are the most used in seabird-environment relationship studies [@tremblay2009], capturing processes at a range of scales from large-scale water-mass properties (e.g. sea surface temperature) to increased predictability of prey availability in mesoscale oceanographic features (e.g. fronts and eddies; @weimerskirch2007; @scales2014). Seabird records were matched with the best open-access, remote-sensed or interpolated data available, and distance from the coast was calculated in R software (see details below).

@@

data.frame(
  Variable = c("Sea Surface Temperature", "Sea Surface Temperature gradient", "Sea Surface Salinity", "Mixed Layer Depth", "8-day average Chlorophyll-a", "Log10 8-day average Chlorophyll-a", "Eddy Kinetic Energy", "Climatic mean EKE", "Climatic standard deviation EKE", "Climatic mean SSTgrad", "Bathymetry", "Slope", "Distance from coast"),
  Abbreviation = c("sst", "sstgrad", "sss", "mld", "chl8", "log10_chl8", "eke", "clim_eke", "clim_ekesd", "clim_sstgrad", "bat", "slope", "dist_coast"),
  Units = c("°Celsius", "delta-°Celsius", "PSU", "m", "mg/m^3", "log10(chl8)", "m/s", "mean(m/s)/season/10 years", "sd(m/s)/season/10 years", "mean(delta-°Celsius)/season/10 years", "m", "degrees", "km"),
  `Spatial resolution` = c("0.25° × 0.25°", "0.25° × 0.25°", "0.083° × 0.083°", "0.083° × 0.083°", "0.036° × 0.036°", "0.036° × 0.036°", "0.2° × 0.2°", "0.2° × 0.2°", "0.2° × 0.2°", "0.25° × 0.25°", "0.004° × 0.004°", "0.004° × 0.004°", "NA"),
  `Temporal resolution` = c("daily", "daily", "daily", "daily", "8-days composite", "8-days composite", "daily", "seasonal climatology", "seasonal climatology", "seasonal climatology", "NA", "NA", "NA"),
  Source = c("Huang et al. (2021)", "Derived from Huang et al. (2021)", "E.U. Copernicus Marine Service Information", "E.U. Copernicus Marine Service Information", "ERDDAP/NOAA", "Derived from ERDDAP/NOAA", "IMOS/AODN", "Derived from IMOS/AODN", "Derived from IMOS/AODN", "Derived from Huang et al. (2021)", "GEBCO", "Derived from GEBCO", "Calculated in R (see main text)")
) |>
  gt::gt() |>
  ## SST & SSTgrad
  gt::tab_footnote(footnote = 
                     gt::md("[NOAA OISST](https://psl.noaa.gov/data/gridded/data.noaa.oisst.v2.highres.html)"),
                   locations = gt::cells_body(columns = Source, rows = c(1:2, 10))) |>
  ## SSS & MLD
  gt::tab_footnote(footnote = 
                     gt::md("2011--2019: [GLOBAL_MULTIYEAR_PHY_001_030](https://doi.org/10.48670/moi-00021); 2020--2021: [GLOBAL_ANALYSIS_FORECAST_PHY_001_024](https://doi.org/10.48670/moi-00016)"),
                   locations = gt::cells_body(columns = Source, rows = 3:4)) |>
  ## CHL8 & log10-CHL8
  gt::tab_footnote(footnote = 
                     gt::md("[ERDDAP/NOAA](https://coastwatch.pfeg.noaa.gov/erddap/griddap/erdMH1chla8day.html)"),
                   locations = gt::cells_body(columns = Source, rows = 5:6)) |>
  ## EKE and derivations
  gt::tab_footnote(footnote = 
                     gt::md("[IMOS/AODN](https://portal.aodn.org.au/search), 'OceanCurrent - Gridded sea level anomaly - Delayed mode' data set"),
                   locations = gt::cells_body(columns = Source, rows = 7:9)) |>
  ## GEBCO & slope
  gt::tab_footnote(footnote = 
                     gt::md("[GEBCO](https://www.gebco.net/data_and_products/gridded_bathymetry_data/)"),
                   locations = gt::cells_body(columns = Source, rows = 11:12)) |>
  ## Improve cols names
  gt::cols_label(
    Variable = gt::md("**Variable**"),
    Abbreviation = gt::md("**Abbreviation**"),
    Units = gt::md("**Units**"),
    `Spatial.resolution` = gt::md("**Spatial resolution**"), 
    `Temporal.resolution` = gt::md("**Temporal resolution**"),
    Source = gt::md("**Source**")) |>
  ## Table title
  gt::tab_header(title = gt::md("**TABLE 1.** Environmental data used in the Region of Common Profile models applied to seabird data off eastern Australia, their spatial/temporal resolutions, units, and sources"))

@@

Data wrangling, modelling and visualization were done in R 4.2.0 [@r2022], using packages **X, Y, ..., Z**. Code and (gridded) data are available in **Daudt et al. 2023 (Zenodo repo)**.

Before data extraction, a few environmental layers needed to be calculated. Spatial and temporal persistence of features such as eddy kinetic energy (EKE) and thermal fronts (SSTgrad) may influence seabird movement and distribution by offering a higher likelihood of finding food [@weimerskirch2007; @scales2014a]. Therefore, for EKE and SSTgrad, we calculated climatic, seasonal average layers based on 10-year data (2011--2021), as well as EKE standard deviation (EKEsd) over the same period. EKE was calculated as $0.5 (u'² + v'²)$, where $u' = u - \bar{u}$ and $v' = v - \bar{v}$ are the surface zonal (latitudinal) and meridian (longitudinal) geostrophic currents, respectively, using a rolling mean of 90 days as $\bar{u}$ and $\bar{v}$ (**ROB: Ref here?**). Sea surface temperature gradient (SSTgrad) was calculated as the spatial gradient by identifying the maximum value for each neighbour cell and averaging them (average neighbourhood maximum), from north-to-south (NS) and west-to-east (WE); then, the magnitude of the gradient was defined as $NS² + WE²$ [@burrows2011]. EKE and SSTgrad calculations were done layer-by-layer to match seabird data on the finest scale possible. The slope was calculated using `raster::terrain()`, and distance from the coast was calculated through `sf::st_distance()` using Australia and Tasmania as mainland (ignoring smaller islands). The most proximate value for each seabird record was extracted from each environmental layer using built-in functions from `raster`, `terra`, and `sf` packages.

>>>>

---
title: "Probabilistic bioregion models suggest seabird assemblages are linked to the major boundary current off eastern Australia"
author:
  - Nicholas W. Daudt:
      email: nicholaswdaudt@gmail.com
      institute: [MarSciUO]
      correspondence: true
  - Eric J. Woehler:
      institute: [ASG]
      correspondence: false
  - Matthew R. Schofield:
      institute: [MathStatUO]
      correspondence: false
  - Robert O. Smith:
      institute: [MarSciUO]
      correspondence: false
  - Leandro Bugoni:
      institute: [FURG]
      correspondence: false
  - William J. Rayment:
      institute: [MarSciUO]
      correspondence: false
institute:
  - MarSciUO: Department of Marine Science, University of Otago, Aotearoa New Zealand
  - ASG: Australasian Seabird Group, BirdLife Australia, Hobart, Australia
  - MathStatUO: Department of Mathematics and Statistics, University of Otago, Aotearoa New Zealand
  - FURG: Institute of Biological Sciences, Universidade Federal do Rio Grande--FURG, Rio Grande, RS, Brazil
abstract: |
  **Aim**: To identify assemblages of seabird species, probabilistically, and relate them to environmental conditions off eastern Australia, a world seabird hotspot.
  **Methods**: We used seabird at-sea ship-based survey data collected over 37° latitude and fitted Region of Common Profile (RCP) mixture models for each season and type of data (presence-absence and abundance). RCP groups are defined as a set of regions where the probability of encountering a specific species profile is constant within regions, but different among them. These groups also vary according to covariates, which in our case included oceanographic, climatic, and physiographic layers.
  **Results**: All models suggested two large scale assemblages ('northern' and 'southern'), except for the autumn presence-absence model which indicated three groups. Model results seem to agree in a biogeographic barrier at ~34°S, where the East Australian Current dissipates as it flows eastward and loses force. Moreover, sea surface temperature or sea surface salinity, were selected in all final models, further indicating a close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and abundance, resulted in similar spatial and species profile patterns.
  **Conclusion**: RCP models clearly identified two seabird assemblages off the east coast of Australia, suggesting the persistence of these groups at seasonal and large spatial scales. Given the rapid change that the East Australian Current is experiencing, and its importance in driving distributions of seabirds, the methods applied in our study could be replicated in future analyses to assess possible changes in seabird assemblages, and how they are affected by changing environmental conditions.
output: 
  # rticles::sim_article:
  #   keep_tex: TRUE
  bookdown::word_document2:
    pandoc_args:
      - --lua-filter=../docs/scholarly-metadata.lua
      - --lua-filter=../docs/author-info-blocks.lua
      - --reference-doc=../docs/docx-template.docx 
bibliography: ../references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
---

@@

```{r setup, include = FALSE, echo = FALSE}
library(readr)
library(dplyr)
library(purrr)
library(gt)
```

@@

Bioregionalization is the discipline that identifies, delimits, characterises, and names biogeographical regions [@vilhena2015; @morrone2018]. Although there is often no agreement on the naming systems between proposed regionalizations [but see @ebach2008], the hierarchical nature of bioregions is well recognized [@vilhena2015; @kreft2010]. Thus, bioregions are usually described within geographically larger classifications, such as 'provinces' or 'realms' [@spalding2007; @costello2017]. The delimitation of bioregions is made through identifying sets of species that are endemic or have high levels of co-occurrence, and are distinct enough from other regions. In this sense, bioregions are interpreted as a natural representations of common historical and/or ecological processes shaped by physical and biological forces [@kreft2010]. In particular, the identification of bioregions allows representation of the distribution of multiple species at a given time and space as well as relating these regions to environmental characteristics. By doing so, bioregions can be interpreted as species assemblages [@lyons2017; @hill2017]. Given these features, bioregions can be used as surrogates for biodiversity monitoring in data-poor regions and in regional conservation planning [@spalding2007; @ferrier2002a], playing a fundamental role in conservation biogeography [@whittaker2005].

Seabird assemblages have been widely described and related to environmental characteristics. Considering the hierarchical nature of oceanographic processes and their resulting features [@haury1978patterns], it is not surprising that the distribution of highly mobile seabirds are linked differently across scales [@hunt1987]. In the marine environment, large scale oceanographic features (e.g. water masses and currents) act as 'biomes' and influence seabird assemblage structure--i.e. the occurrence of species and their relative numerical composition [@ribic1997; @hyrenbach2007; @commins2014]. In contrast, local abundance is related to temporary processes that aggregate prey and therefore predators, such as fronts and eddies [@weimerskirch2007; @scales2014]. Such smaller-scale processes, although ephemeral, are primarily driven by physical and atmospheric forces, and are thus somewhat predictable in space and time, providing consistent feeding areas for marine predators when they occur [@davoren2013; @hazen2013]. Noticeably, sea surface temperature is a key driver of seabird distribution, at individual- and assemblage-levels, operating over a range of scales from large (e.g. water masses) to small (e.g. fronts) [@dunlop1988; @ribic1997; @quillfeldt2015]. Nonetheless, many regions still lack information regarding distribution of seabird species and how they assemble in space and time.

One area that could benefit from better understanding of seabird distributions is Australia, considered a world seabird hotspot [@karpouzi2007; @ramírez2017]. Given the region's importance for seabird diversity, there are surprisingly few studies on seabird distribution [@mott2018; @bernard2021]. Off northwest Australia, at-sea surveys, multivariate analyses [@dunlop1988] and ensemble species distribution models of species assemblages [@lavers2014] have shown that habitat use by tropical species is related to sea surface temperature and salinity. In particular, @dunlop1988 suggested that salinity is the main variable influencing species distribution in oceanic waters whereas temperature is the key factor on the continental shelf. In contrast, there are few studies at large scales off eastern Australia. @reid2000 and @evans2021a analysed seabird assemblages and their relationships with oceanography and prey data at broad scales (tens of km), off southeast Tasmania. The remaining studies are mostly descriptive accounts at local scales [e.g. @norman1992distribution; @blaber1994distribution], or focused on a single species [e.g. @priddel2014].

The East Australian Current (EAC) is the major western boundary current in the South Pacific, and the dominant current in eastern Australia [@ganachaud2014]. The EAC has been extensively studied, although its link between oceanography and biological processes is still in its infancy. The dynamic features caused by the EAC, such as eddies and upwellings, are likely to influence marine biota [@suthers2011; @scales2014]. While most of the biological studies done in this system were focused on plankton entrapment within the cores of eddies [@garcia2022; @firme2023; @hassler2011], there is a growing body of work focusing on fisheries [e.g. @hobday2010; @young2011; @brieva2015]. These studies emphasize how the probability of occurrence of a given taxon, and the structure of species assemblages, is driven by the EAC. Notwithstanding, the relationship between oceanography and marine megafauna along the EAC is poorly known [@suthers2011]. Some progress has been made using shark and ray movement data [@jaine2014; @niella2020; @niella2022], and satellite tracking has also demonstrated that increased sea surface temperature in EAC decreases penguin foraging success [@carroll2016]. However, information at the assemblage-level is still scarce [but see @evans2021a].

Marine species range shifts are more likely to occur faster than terrestrial species due to their higher sensitivity to warming [@pinsky2019; @lenoir2020]. As a result, climate change is rearranging the structure of species assemblages around the globe [@poloczanska2013; @antão2020; @dornelas2014]. Top predators, for instance, may change their core habitats [@hazen2013a; @reisinger2022], with unprecedented population consequences. For seabirds, @péron2010 showed that sea surface temperature anomalies affected the occurrence and abundance of albatrosses and petrels species along a latitudinal range in the Indian Ocean, thus affecting assemblage structure. In eastern Australia, changing EAC conditions have affected a range of species. Changes in zooplankton assemblage structure and fish distribution patterns have been recorded, and shark distribution shifts are expected to occur [@johnson2011; @niella2022]. One of the strongest marine heatwave ever recorded, in the Tasman Sea, showed drastic environmental impacts [@oliver2017]. Therefore, given that the EAC is strengthening and getting warmer [@cai2005; @phillips2022], biological changes are expected to occur [@poloczanska2007; @hobday2010].

Given east Australia's importance for seabirds and the recent attention on how climate change in the EAC can affect marine biota [@poloczanska2007], an assessment of the spatial distribution of seabirds at large scales is timely. Seabirds can serve as sentinels for tracking ecological changes in the EAC in many ways [e.g. @sydeman2012]. Surprisingly, though, seabird distribution is under-represented in climate change studies, despite being highlighted as a priority topic [@grémillet2009; @sydeman2012]. In this study, we aim to identify assemblages of seabird species probabilistically and relate them to environmental conditions, by using at-sea surveys data and a state-of-the-art modelling technique. We used the mixture-model Region of Common Profiles [RCP; @foster2013], regarded as one of the best available tools for describing bioregions [@hill2020; @woolley2020]. RCP is a model-based approach, and has a range of advantages over traditional algorithm-based statistics [@warton2015; @hill2020]. East Australia supports a high diversity of seabirds and offers wide environmental gradients, from the tropics to the subantarctic. Therefore, we expect to find different assemblages relating to different environmental conditions. In addition, we run RCP models for presence-absence and abundance as input data, to test if the different numeric resolution would result in similar assemblage patterns. Furthermore, we provide a repeatable framework on which future studies can build to assess how changes in the EAC will influence the distribution of seabird assemblages.

@@

The EAC originates from the South Equatorial Current (SEC), a tropical, nutrient-poor water. When the SEC's North Caledonian Jet hits the Queensland Plateau (\~18°S), it splits into a north branch running towards the Gulf of Papua and a south branch, which marks the start of the EAC [@kessler2013; @ganachaud2014]. The South Caledonian Jet completes the EAC system at \~22°S after encountering the shelf break in the southern portion of the Great Barrier Reef (**Fig. 1a**; @kessler2013; @ganachaud2014). **SST/SSS characteristics? ROB?** The EAC interacts with the topography of the seafloor, giving rise to oceanographic features and processes, such as smaller currents, upwellings, and persistent dynamic eddies [@oke2000; @ridgway2003a; @oke2019]. The EAC is mainly attached to the shelf break, but at \~32°S it spawns an area of eddy generation that dissipates throughout the western Tasman Sea [@mata2006; @oke2019], the 'EAC eastern extension' (@oke2019a; often termed the 'Tasman Front'). Although most EAC energy is converted into eddies, the EAC still runs southwards following the shelf break towards Tasmania (the 'EAC southern extension'; **Fig. 1a**), possibly carrying eddy-like features as a transient current [@oke2019].

@@

Seabird data were gathered aboard RV Investigator and RV Falkor during 15 voyages, between 2016--2021, over \~37° of latitude (**Fig. 1b**; **Table S1**). One observer continuously recorded seabirds within a 300-m forward quadrant (eye height **18-m, Eric?** a.s.l) throughout daylight hours while the ship was underway at a constant speed. The species, the number of birds per group, and behaviour were logged directly into a data entry system through a ship-based web server. Individual records received a date/time/geographic coordinate label (multi-species records have the same label but are logged as separate entries for each species) **(Eric: is there any ref for the protocol?)**. Birds following the boat were previously excluded from analyses.

```{r fig1-study-area, fig.cap = "**FIGURE 1.** Study area showing the main ocean currents (a), and seabird records made by season on top of the 1° latitude × 1° longitude grid cells (b). In (a), the East Australian Current (EAC) system is highlighted with its acronyms in bold and drawn in blue. The South Equatorial Current (SEC) and its branches, the North Caledonian Jet (NCJ) and South Caledonian Jet (SCJ), are drawn in red. The EAC 'eastern extension' is an eddy field (blue shaded area) often called the 'Tasman Front', from which the East Auckland Current originates (EAUC, in green). Figure (a) is adapted after Oke, Roughan, et al. (2019); for colour references see the online version.", fig.align = "center", echo = FALSE}

@@

To explain and predict species distribution and biogeography, a thoughtful choice of covariates is important [@barbet-massin2014]. We used three physiographic variables and seven oceanographic variables, from which a further three climatic layers were calculated (**Table 1**). These variables are the most used in seabird-environment relationship studies [@tremblay2009], capturing processes at a range of scales from large-scale water-mass properties (e.g. sea surface temperature) to prey availability in mesoscale oceanographic features (e.g. fronts and eddies; @weimerskirch2007; @scales2014). Seabird records were matched with the best open-access, remote-sensed or interpolated data available, and distance from the coast was calculated in R software, version 4.2.0 [@r2022] (see details below).

@@

data.frame(
  Variable = c("Sea Surface Temperature", "Sea Surface Temperature gradient", "Sea Surface Salinity", "Mixed Layer Depth", "8-day average Chlorophyll-a", "Log10 8-day average Chlorophyll-a", "Eddy Kinetic Energy", "Climatic mean EKE", "Climatic standard deviation EKE", "Climatic mean SSTgrad", "Bathymetry", "Slope", "Distance from coast"),
  Abbreviation = c("sst", "sst_grad", "sss", "mld", "chl8", "log10_chl8", "eke", "clim_eke_mean", "clim_eke_sd", "clim_sst_grad", "bat", "slope", "dist_coast"),
  Units = c("°Celsius", "delta-°Celsius", "PSU", "m", "mg/m^3", "log10(chl8)", "m/s", "mean(m/s)/season/10 years", "sd(m/s)/season/10 years", "mean(delta-°Celsius)/season/10 years", "m", "degrees", "km"),
  `Spatial resolution` = c("0.25° × 0.25°", "0.25° × 0.25°", "0.083° × 0.083°", "0.083° × 0.083°", "0.036° × 0.036°", "0.036° × 0.036°", "0.2° × 0.2°", "0.2° × 0.2°", "0.2° × 0.2°", "0.25° × 0.25°", "0.004° × 0.004°", "0.004° × 0.004°", "NA"),
  `Temporal resolution` = c("daily", "daily", "daily", "daily", "8-days composite", "8-days composite", "daily", "seasonal climatology", "seasonal climatology", "seasonal climatology", "NA", "NA", "NA"),
  Source = c("Huang et al. (2021)", "Derived from Huang et al. (2021)", "E.U. Copernicus Marine Service Information", "E.U. Copernicus Marine Service Information", "ERDDAP/NOAA", "Derived from ERDDAP/NOAA", "IMOS/AODN", "Derived from IMOS/AODN", "Derived from IMOS/AODN", "Derived from Huang et al. (2021)", "GEBCO", "Derived from GEBCO", "Calculated in R (see main text)")
) |>
  gt::gt() |>
  ## SST & SSTgrad
  gt::tab_footnote(footnote = 
                     gt::md("[NOAA OISST](https://psl.noaa.gov/data/gridded/data.noaa.oisst.v2.highres.html)"),
                   locations = gt::cells_body(columns = Source, rows = c(1:2, 10))) |>
  ## SSS & MLD
  gt::tab_footnote(footnote = 
                     gt::md("2011--2019: [GLOBAL_MULTIYEAR_PHY_001_030](https://doi.org/10.48670/moi-00021); 2020--2021: [GLOBAL_ANALYSIS_FORECAST_PHY_001_024](https://doi.org/10.48670/moi-00016)"),
                   locations = gt::cells_body(columns = Source, rows = 3:4)) |>
  ## CHL8 & log10-CHL8
  gt::tab_footnote(footnote = 
                     gt::md("[ERDDAP/NOAA](https://coastwatch.pfeg.noaa.gov/erddap/griddap/erdMH1chla8day.html)"),
                   locations = gt::cells_body(columns = Source, rows = 5:6)) |>
  ## EKE and derivations
  gt::tab_footnote(footnote = 
                     gt::md("[IMOS/AODN](https://portal.aodn.org.au/search), 'OceanCurrent - Gridded sea level anomaly - Delayed mode' data set"),
                   locations = gt::cells_body(columns = Source, rows = 7:9)) |>
  ## GEBCO & slope
  gt::tab_footnote(footnote = 
                     gt::md("[GEBCO](https://www.gebco.net/data_and_products/gridded_bathymetry_data/)"),
                   locations = gt::cells_body(columns = Source, rows = 11:12)) |>
  ## Improve cols names
  gt::cols_label(
    Variable = gt::md("**Variable**"),
    Abbreviation = gt::md("**Abbreviation**"),
    Units = gt::md("**Units**"),
    `Spatial.resolution` = gt::md("**Spatial resolution**"), 
    `Temporal.resolution` = gt::md("**Temporal resolution**"),
    Source = gt::md("**Source**")) |>
  ## Table title
  gt::tab_header(title = gt::md("**TABLE 1.** Environmental data used in the Region of Common Profile models applied to seabird data off eastern Australia, their spatial/temporal resolutions, units, and sources."))

@@

Data wrangling, modelling and visualization were done in R 4.2.0 [@r2022], using packages 'plyr' 1.8.8 [@plyr], 'dplyr' 1.1.2 [@dplyr], 'tidyr' 1.3.0 [@tidyr], 'readr' 2.1.4 [@readr], 'tibble' 3.2.1 [@tibble], 'lubridate' 1.9.2 [@lubridate], 'stringr' 1.5.0 [@stringr], 'purrr' 1.0.1 [@purrr], 'ggplot2' 3.4.2 [@ggplot2], 'ggspatial' 1.1.7 [@ggspatial], 'patchwork' 1.1.2 [@patchwork], 'RColorBrewer' 1.1-3 [@rcolorbrewer], 'rnaturalearth' 0.3.2 [@rnaturalearth], 'sp' 1.6-0 [@sp1; @sp2], 'sf' 1.0-8 [@sf], 'mapview' 2.11.0 [@mapview], 'raster' 3.5-21 [@raster], 'terra' 1.6-7 [@terra], 'rerddap' 1.0.2 [@rerddap], 'rerddapXtracto' 1.1.4 [@rerddapXtracto], 'hadsstR' [@hadsstr], 'corrplot' 0.92 [@corrplot2021], 'iNEXT' 3.0.0 [@chao2014; @hsieh2016], and 'ecomix' 1.0.0 [@ecomix]. Code and the modelling dataset (see \@ref(modellingdatasets)) are available in **Daudt et al. 2023 (Zenodo repo)**.

### Environmental layers
